---
title: "C42B cells day1 and 5, comparing Ola 1 and 5."
author: "Marion Hardy"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true 
    theme: spacelab 
    highlight: monochrome
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, cache = TRUE, echo = FALSE, warning = F, cache.lazy = F)
knitr::opts_chunk$set(fig.width=6, fig.height=7) 

library(RColorBrewer)
library(tidyverse)
library(DESeq2)
library(cowplot)
library(dendextend)
library(usedist)
library(biomaRt)
library(ggrepel)
library(pheatmap)
library(ComplexHeatmap)
library(xlsx)
library(msigdbr)
library(clusterProfiler)
library(kableExtra)

ensembl_to_geneName <- readRDS("./data/hsapiens_annotations_230510.rds") %>% 
  filter(!duplicated(gene, drop = F))

dir = getwd()

```

# Introduction

BulkRNAseq analysis of C42B cells treated with olaparib at different
concentrations (1-5 uM?) for different amount of times (1-9 days).

This report will present a broad overview of how similar/different
samples are in terms of gene expression. The objective is to help AK
narrow down which gene sets are of interest and follow this with a gsea
analysis and a closer look at the function of these identified gene
groups.

Each sample comparison shown in this report will have additional files
generated:

-   res_tbl.csv = all differentially expressed genes, annotated (no
    filtering)

-   Sign_genes.xlsx = excel doc with all significantly up- and
    down-regulated genes (log2Fc \>= 1), annotated and sorted into all
    genes, only up and only down on different sheets.

-   volcano plot of all genes (html format)

NB: Data quality has been checked within the previous script "DESeq2_1-18.R",
everything looked good.

```{r}

dds1 = readRDS("./data_output/C42B_DESeq_1-18_Day1_vs_DMSO.Rds")
dds5 = readRDS("./data_output/C42B_DESeq_1-18_Day5_vs_DMSO.Rds")

```


```{r}

# MsigDb

hsa_GOBP_sets = msigdbr(
  species = "Homo sapiens", 
  category = "C5",
  subcategory = "GO:BP")

hsa_GOMF_sets = msigdbr(
  species = "Homo sapiens", 
  category = "C5",
  subcategory = "GO:MF")

hsa_reactome_sets = msigdbr(
  species = "Homo sapiens", 
  category = "C2",
  subcategory = "CP:REACTOME") # for reactome collection

set.seed(054057) # for reproducibility

```

# PCA

```{r, fig.height=5, fig.width=6}
rld = vst(dds1)

plotPCA(rld,intgroup="Combined") + 
  geom_label_repel(aes(label = Combined), 
                   size = 3, box.padding = 0.6,
                   max.overlaps = Inf)+
  xlim(c(-20,20))+
  ylim(c(-15, 15))+
  theme_bw()+
  labs(title = 'PCA per treatment and group')
```

# Euclidian distances

```{r, fig.height=5, fig.width=6}
sampleDists = dist(t(assay(rld)))
sampleDistMatrix = as.matrix(sampleDists)
rownames(sampleDistMatrix) = paste(rld$Combined, sep="-")
colnames(sampleDistMatrix) = paste(rld$Combined, sep="-")
colors = colorRampPalette(rev(brewer.pal(9, "Blues")) )(255)

pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
```

# Dendogram

```{r, fig.height=4, fig.width=7}

sampleDists_named = sampleDists
sampleDists_named = dist_setNames(sampleDists_named, as.vector(colData(dds1)[,8]))

hc = hclust(sampleDists_named)
hc %>% 
  as.dendrogram() %>% 
  set("labels_col", c(1,2,4),k=3) %>% 
  plot()


```


# Day1 Ola 1 and 5 vs DMSO

## Volcano plots

```{r, fig.show="hold", out.width="50%"}

for (i in 2:3){
    cond = resultsNames(dds1)[i]
    name = gsub("Treatment_","",resultsNames(dds1)[i])
    
    res <- results(dds1, name = cond)
    res_tbl <- as_tibble(res, rownames="ensembl")
    res_tbl <- res_tbl %>%
    left_join(ensembl_to_geneName) %>%
    arrange(padj) 
    dir.create(paste0(dir,'/data_output/',name))
    dir.create(paste0(dir,'/figures/',name))
    write.csv(res_tbl, paste0("./data_output/",name,"/res_tbl.csv"), row.names = T)
    
    overexpr <- res_tbl %>% filter(padj<=0.05 & log2FoldChange>=1) 
    underexpr <- res_tbl %>% filter(padj<=0.05 & log2FoldChange<=-1)
    signif <- full_join(overexpr, underexpr)
  
    write.xlsx(overexpr, file= paste0("./data_output/",name,"/Sign_genes.xlsx"), sheetName="UP")
    write.xlsx(underexpr, file=paste0("./data_output/",name,"/Sign_genes.xlsx"), sheetName="DOWN", append=TRUE)
    write.xlsx(signif, file=paste0("./data_output/",name,"/Sign_genes.xlsx"), sheetName="SIGNIF", append=TRUE)
    
    p1 =
  res_tbl %>%
    filter(!is.na(padj)) %>%
    ggplot(aes(x = log2FoldChange, y = -log10(padj),
               color = padj < 0.05 & abs(log2FoldChange) > 1, 
               label = ifelse(padj<0.05&log2FoldChange>=1|
                                padj<0.05&log2FoldChange<=-1,as.character(gene),''))) +
    scale_colour_manual(values = c("gray", "firebrick3")) +
    labs(col="Significantly expressed")+
    geom_point(size = 0.5) +
    geom_hline(yintercept = -log10(0.05)) +
    geom_vline(xintercept = 1) +
    geom_vline(xintercept = -1)+
    geom_text_repel(max.overlaps = 15,
                    box.padding = 0.25,
                    segment.color = 'grey50',
                    fontface = "italic")+
    labs(title = name)+
    theme_bw()
  
  print(p1)
  
    ggsave(paste0("./figures/",name,"/volcanoplot.svg"), last_plot(), 
           dpi= 500, width = 12, height = 8)
  
}

```

## GSEA Day1 Ola1
### GOBP

```{r, fig.height=5, fig.width=7}

res_tbl = read.csv(paste0("./data_output/Combined_Day1_Ola1_vs_Day1_DMSO/res_tbl.csv"))
    
ordered_genes_fc = res_tbl$log2FoldChange
names(ordered_genes_fc) = res_tbl$gene
ordered_genes_fc = sort(ordered_genes_fc, decreasing = T)

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_GOBP_sets,
    gs_name,
     gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 6)+ 
  labs(title = "GSEA",
          subtitle = paste0("GSEA GO:BP Day 1 Ola 1 vs DMSO"))+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

ggsave(plot = last_plot(), paste0("./figures/Combined_Day1_Ola1_vs_Day1_DMSO/GOBPgsea.svg"), dpi = 300,
      height = 8, width = 8, device = "svg")

write.xlsx(gsea_results, paste0("./data_output/Combined_Day1_Ola1_vs_Day1_DMSO/GOBPgsea.xlsx"))

gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```

### Reactome

no results


## GSEA Day1 Ola5
### GOBP

```{r}

res_tbl = read.csv(paste0("./data_output/Combined_Day1_Ola5_vs_Day1_DMSO/res_tbl.csv"))
    
ordered_genes_fc = res_tbl$log2FoldChange
names(ordered_genes_fc) = res_tbl$gene
ordered_genes_fc = sort(ordered_genes_fc, decreasing = T)

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_GOBP_sets,
    gs_name,
     gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 6)+ 
  labs(title = "GSEA",
          subtitle = paste0("GSEA GO:BP Day 1 Ola 5 vs DMSO"))+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

ggsave(plot = last_plot(), paste0("./figures/Combined_Day1_Ola5_vs_Day1_DMSO/GOBPgsea.svg"), dpi = 300,
      height = 8, width = 8, device = "svg")

write.xlsx(gsea_results, paste0("./data_output/Combined_Day1_Ola5_vs_Day1_DMSO/GOBPgsea.xlsx"))

gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```

### Reactome

```{r}

res_tbl = read.csv(paste0("./data_output/Combined_Day1_Ola5_vs_Day1_DMSO/res_tbl.csv"))
    
ordered_genes_fc = res_tbl$log2FoldChange
names(ordered_genes_fc) = res_tbl$gene
ordered_genes_fc = sort(ordered_genes_fc, decreasing = T)

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_reactome_sets,
    gs_name,
     gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 6)+ 
  labs(title = "GSEA",
          subtitle = paste0("GSEA Reactome Day 1 Ola 5 vs DMSO"))+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

ggsave(plot = last_plot(), paste0("./figures/Combined_Day1_Ola5_vs_Day1_DMSO/Reactomegsea.svg"), dpi = 300,
      height = 8, width = 8, device = "svg")

write.xlsx(gsea_results, paste0("./data_output/Combined_Day1_Ola5_vs_Day1_DMSO/Reactomegsea.xlsx"))

gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```


# Day 3 Ola 1 and 5 vs DMSO

## Volcano plots

```{r, fig.show="hold", out.width="50%"}

for (i in c(5:6)){
    cond = resultsNames(dds5)[i]
    name = gsub("Treatment_","",resultsNames(dds5)[i])
    
    res <- results(dds5, name = cond)
    res_tbl <- as_tibble(res, rownames="ensembl")
    res_tbl <- res_tbl %>%
    left_join(ensembl_to_geneName) %>%
    arrange(padj) 
    dir.create(paste0(dir,'/data_output/',name))
    dir.create(paste0(dir,'/figures/',name))
    write.csv(res_tbl, paste0("./data_output/",name,"/res_tbl.csv"), row.names = T)
    overexpr <- res_tbl %>% filter(padj<=0.05 & log2FoldChange>=1) 
    underexpr <- res_tbl %>% filter(padj<=0.05 & log2FoldChange<=-1)
    signif <- full_join(overexpr, underexpr)
  
    write.xlsx(overexpr, file= paste0("./data_output/",name,"/Sign_genes.xlsx"), sheetName="UP")
    write.xlsx(underexpr, file=paste0("./data_output/",name,"/Sign_genes.xlsx"), sheetName="DOWN", append=TRUE)
    write.xlsx(signif, file=paste0("./data_output/",name,"/Sign_genes.xlsx"), sheetName="SIGNIF", append=TRUE)
    
    p1 =
  res_tbl %>%
    filter(!is.na(padj)) %>%
    ggplot(aes(x = log2FoldChange, y = -log10(padj),
               color = padj < 0.05 & abs(log2FoldChange) > 1, 
               label = ifelse(padj<0.05&log2FoldChange>=1|
                                padj<0.05&log2FoldChange<=-1,as.character(gene),''))) +
    scale_colour_manual(values = c("gray", "firebrick3")) +
    labs(col="Significantly expressed")+
    geom_point(size = 0.5) +
    geom_hline(yintercept = -log10(0.05)) +
    geom_vline(xintercept = 1) +
    geom_vline(xintercept = -1)+
    geom_text_repel(max.overlaps = 15,
                    box.padding = 0.25,
                    segment.color = 'grey50',
                    fontface = "italic")+
    labs(title = name)+
    theme_bw()
  
  print(p1)
  
    ggsave(paste0("./figures/",name,"/volcanoplot.svg"), last_plot(), 
           dpi= 500, width = 12, height = 8)
  
    
}


```


## GSEA Day5 Ola1
### GOBP

```{r}

res_tbl = read.csv(paste0("./data_output/Combined_Day5_Ola1_vs_Day5_DMSO/res_tbl.csv"))
    
ordered_genes_fc = res_tbl$log2FoldChange
names(ordered_genes_fc) = res_tbl$gene
ordered_genes_fc = sort(ordered_genes_fc, decreasing = T)

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_GOBP_sets,
    gs_name,
     gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 6)+ 
  labs(title = "GSEA",
          subtitle = paste0("GSEA GO:BP Day 5 Ola 1 vs DMSO"))+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

ggsave(plot = last_plot(), paste0("./figures/Combined_Day5_Ola1_vs_Day5_DMSO/GOBPgsea.svg"), dpi = 300,
      height = 8, width = 8, device = "svg")

write.xlsx(gsea_results, paste0("./data_output/Combined_Day5_Ola1_vs_Day5_DMSO/GOBPgsea.xlsx"))

gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```

### Reactome

```{r}

res_tbl = read.csv(paste0("./data_output/Combined_Day5_Ola1_vs_Day5_DMSO/res_tbl.csv"))
    
ordered_genes_fc = res_tbl$log2FoldChange
names(ordered_genes_fc) = res_tbl$gene
ordered_genes_fc = sort(ordered_genes_fc, decreasing = T)

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_reactome_sets,
    gs_name,
     gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 6)+ 
  labs(title = "GSEA",
          subtitle = paste0("GSEA Reactome Day 5 Ola 1 vs DMSO"))+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

ggsave(plot = last_plot(), paste0("./figures/Combined_Day5_Ola1_vs_Day5_DMSO/Reactomegsea.svg"), dpi = 300,
      height = 8, width = 8, device = "svg")

write.xlsx(gsea_results, paste0("./data_output/Combined_Day5_Ola1_vs_Day5_DMSO/Reactomegsea.xlsx"))

gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```


## GSEA Day5 Ola5
### GOBP

```{r}

res_tbl = read.csv(paste0("./data_output/Combined_Day5_Ola5_vs_Day5_DMSO/res_tbl.csv"))
    
ordered_genes_fc = res_tbl$log2FoldChange
names(ordered_genes_fc) = res_tbl$gene
ordered_genes_fc = sort(ordered_genes_fc, decreasing = T)

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_GOBP_sets,
    gs_name,
     gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 6)+ 
  labs(title = "GSEA",
          subtitle = paste0("GSEA GO:BP Day 5 Ola 5 vs DMSO"))+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

ggsave(plot = last_plot(), paste0("./figures/Combined_Day5_Ola5_vs_Day5_DMSO/GOBPgsea.svg"), dpi = 300,
      height = 8, width = 8, device = "svg")

write.xlsx(gsea_results, paste0("./data_output/Combined_Day5_Ola5_vs_Day5_DMSO/GOBPgsea.xlsx"))

gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```

### Reactome

```{r}

res_tbl = read.csv(paste0("./data_output/Combined_Day5_Ola5_vs_Day5_DMSO/res_tbl.csv"))
    
ordered_genes_fc = res_tbl$log2FoldChange
names(ordered_genes_fc) = res_tbl$gene
ordered_genes_fc = sort(ordered_genes_fc, decreasing = T)

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_reactome_sets,
    gs_name,
     gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 6)+ 
  labs(title = "GSEA",
          subtitle = paste0("GSEA Reactome Day 5 Ola 5 vs DMSO"))+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

ggsave(plot = last_plot(), paste0("./figures/Combined_Day5_Ola5_vs_Day5_DMSO/Reactomegsea.svg"), dpi = 300,
      height = 8, width = 8, device = "svg")

write.xlsx(gsea_results, paste0("./data_output/Combined_Day5_Ola5_vs_Day5_DMSO/Reactomegsea.xlsx"))

gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```



# Venn diagram on steroids (UpsetPlot)

Conditions we want to compare, per group:

\- Ola 1 vs DMSO

\- Ola 5 vs DMSO

I extracted all significantly differentially expressed genes (DEGs) for
each condition (\|log2Fc\| \>= 1 and pval \< 0.05).

**Set size** = how many DEGs there are per sample

**Intersection size** = How many genes are shared between the x samples
tied together with the bar + nodes

So if you sum up all the intersection size with a dot in a sample, you
get the set size of that sample. It's essentially a venn diagram for
when you have \>3 samples you want to compare.

```{r, fig.height=7, fig.width=10}

# load in the data to be formatted for an upset plot
# keep each gene name that has a signif pval and |log2fc| > 0.5
# use ensembl to encompass unnamed transcripts

folder = c("Combined_Day1_Ola1_vs_Day1_DMSO","Combined_Day1_Ola5_vs_Day1_DMSO",
           "Combined_Day5_Ola1_vs_Day5_DMSO","Combined_Day5_Ola5_vs_Day5_DMSO")

for(i in 1:length(folder)){
  res = read.csv(paste0("./data_output/",folder[i],"/res_tbl.csv"))
  genes = res %>% 
    select(ensembl, log2FoldChange, padj) %>% 
    filter(abs(log2FoldChange)>= 0.5 & padj < 0.05) %>% 
    select(ensembl)
  genes$temp = 1
  colnames(genes)[2] = folder[i]
  assign(folder[i], genes)
}

datalist_dmso = list(Combined_Day1_Ola1_vs_Day1_DMSO,Combined_Day1_Ola5_vs_Day1_DMSO,
           Combined_Day5_Ola1_vs_Day5_DMSO,Combined_Day5_Ola5_vs_Day5_DMSO)

big_data_dmso =
  datalist_dmso %>% purrr::reduce(full_join, by='ensembl')%>% 
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) 

```

```{r, fig.height=5, fig.width=10}

m1 = make_comb_mat(big_data_dmso)

UpSet(m1, top_annotation = upset_top_annotation(m1, add_numbers = TRUE),
    right_annotation = upset_right_annotation(m1, add_numbers = TRUE),
    comb_order = order(-comb_size(m1)), 
    comb_col = c("#ffbe4f", "#e8702a", "#6bd2db","#7e073d")[comb_degree(m1)],
    row_names_gp = gpar(fontsize = 10))


```

