---
title: "C42B cells day1 and 5, comparing Ola 1 and 5."
author: "Marion Hardy"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: spacelab
    highlight: monochrome
  word_document:
    toc: true
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, cache = T, echo = FALSE, warning = F, cache.lazy = F,
                      always_allow_html = T)
knitr::opts_chunk$set(fig.width=6, fig.height=7) 

library(RColorBrewer)
library(viridis)
library(tidyverse)
library(DESeq2)
library(cowplot)
library(dendextend)
library(usedist)
library(biomaRt)
library(ggrepel)
library(pheatmap)
library(ComplexHeatmap)
library(xlsx)
library(msigdbr)
library(clusterProfiler)
library(kableExtra)
library(gridExtra)
library(AMR)
library(enrichplot)

ensembl_to_geneName <- readRDS("./data/hsapiens_annotations_230510.rds") %>% 
  filter(!duplicated(gene, drop = F))

dir = getwd()

```

# Introduction

BulkRNAseq analysis of C42B cells treated with olaparib at different
concentrations (1-5 uM?) for different amount of times (1-9 days).

This report will present a broad overview of how similar/different
samples are in terms of gene expression. The objective is to help AK
narrow down which gene sets are of interest and follow this with a gsea
analysis and a closer look at the function of these identified gene
groups.

Each sample comparison shown in this report will have additional files
generated:

-   res_tbl.csv = all differentially expressed genes, annotated (no
    filtering)

-   Sign_genes.xlsx = excel doc with all significantly up- and
    down-regulated genes (log2Fc \>= 1), annotated and sorted into all
    genes, only up and only down on different sheets.

-   volcano plot of all genes (html format)

NB: Data quality has been checked within the previous script "DESeq2_1-18.R",
everything looked good.

```{r}

dds1 = readRDS("./data_output/C42B_DESeq_1-18_Day1_vs_DMSO.rds")
dds5 = readRDS("./data_output/C42B_DESeq_1-18_Day5_vs_DMSO.rds")

```


```{r}

# MsigDb

hsa_GOBP_sets = msigdbr(
  species = "Homo sapiens", 
  category = "C5",
  subcategory = "GO:BP")

hsa_hallmark_sets = msigdbr(
  species = "Homo sapiens", 
  category = "H")

hsa_reactome_sets = msigdbr(
  species = "Homo sapiens", 
  category = "C2",
  subcategory = "CP:REACTOME") # for reactome collection

set.seed(054057) # for reproducibility

```

# PCA

```{r, fig.height=5, fig.width=6}
rld = vst(dds1)

plotPCA(rld,intgroup="Combined") + 
  geom_text_repel(aes(label = Combined, size = 8), 
                  colour = 'black',max.overlaps = Inf, box.padding = 0.6)+
  xlim(c(-20,40))+
  ylim(c(-17, 17))+
  theme_bw()+
  labs(title = 'PCA per treatment and group')

ggsave(plot = last_plot(),"./figures/for_pub/PCA1.svg", dpi = 300,
       height = 6, width = 7.5)

```

# Extract the weights per genes

```{r, fig.height=5, fig.width=6, echo=FALSE}

data = assay(dds1)
forlabel = as.data.frame(rownames(data))
colnames(forlabel) = "ensembl"
forlabel = left_join(forlabel, ensembl_to_geneName)
rownames(data) = forlabel$gene

pca = prcomp(data)
loadings = as.data.frame(pca$x)
loadings$gene = rownames(loadings)

top_genes = loadings %>% 
  select(gene, PC1, PC2) %>%
  pivot_longer(matches("PC"), names_to = "PC", values_to = "loading") %>% 
  group_by(PC) %>% 
  arrange(desc(abs(loading))) %>% 
  slice(1:20) %>% # take the 20 top rows
  pull(gene) %>% 
  unique()


```

```{r, fig.height=5, fig.width=6, echo=FALSE}

# plot the biplot

loadings %>% 
  filter(gene %in% top_genes) %>% 
  ggplot() +
  geom_segment(aes(x = 0, y = 0, xend = PC1, yend = PC2), 
               arrow = arrow(length = unit(0.1, "in")),
               color = "firebrick3") +
  geom_text(aes(x = PC1, y = PC2, label = gene),
            nudge_y = 10000, nudge_x = -50000,
            size = 3)+
  theme_bw()+
  xlab("PC1 loadings")+
  ylab("PC2 loadings")+
  labs(title = 'Biplot of gene loadings')

ggsave(plot = last_plot(),"./figures/for_pub/PCA1_weights.svg", dpi = 300,
       height = 4.5, width = 5.5)


```

# Euclidian distances

```{r, fig.height=5, fig.width=6}
sampleDists = dist(t(assay(rld)))
sampleDistMatrix = as.matrix(sampleDists)
rownames(sampleDistMatrix) = paste(rld$Combined, sep="-")
colnames(sampleDistMatrix) = paste(rld$Combined, sep="-")
colors = colorRampPalette(rev(brewer.pal(9, "Blues")) )(500)
colors = viridis(20)
# colors = colorRampPalette(rev(brewer.pal(7, "YlGnBu")) )(4)

pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
# figure exported as svg manually
```

# Dendogram

```{r, fig.height=4, fig.width=7}

sampleDists_named = sampleDists
sampleDists_named = dist_setNames(sampleDists_named, as.vector(colData(dds1)[,8]))

hc = hclust(sampleDists_named)
hc %>% 
  as.dendrogram() %>% 
  set("labels_col", c(1,2,4),k=3) %>% 
  plot()
# figure exported as svg manually

```


# Day1 Ola 1 and 5 vs DMSO

## Volcano plots

```{r, fig.show="hold", out.width="50%"}

for (i in 2:3){
    cond = resultsNames(dds1)[i]
    name = gsub("Treatment_","",resultsNames(dds1)[i])
    
    res <- results(dds1, name = cond)
    res_tbl <- as_tibble(res, rownames="ensembl")
    res_tbl <- res_tbl %>%
    left_join(ensembl_to_geneName) %>%
    arrange(padj) 
    dir.create(paste0(dir,'/data_output/',name))
    dir.create(paste0(dir,'/figures/',name))
    write.csv(res_tbl, paste0("./data_output/",name,"/res_tbl.csv"), row.names = T)
    
    overexpr <- res_tbl %>% filter(padj<=0.05 & log2FoldChange>=0.5) 
    underexpr <- res_tbl %>% filter(padj<=0.05 & log2FoldChange<=-0.5)
    signif <- full_join(overexpr, underexpr)
  
    write.xlsx(overexpr, file= paste0("./data_output/",name,"/Sign_genes.xlsx"), sheetName="UP")
    write.xlsx(underexpr, file=paste0("./data_output/",name,"/Sign_genes.xlsx"), sheetName="DOWN", append=TRUE)
    write.xlsx(signif, file=paste0("./data_output/",name,"/Sign_genes.xlsx"), sheetName="SIGNIF", append=TRUE)
    
    top15 = 
      signif %>% 
      arrange(desc(abs(log2FoldChange))) %>% 
      filter(!is.na(gene))
    
    top15 = top15$gene[1:15]
    
    res_tbl <- res_tbl %>%
    filter(!is.na(padj)) %>%
    mutate(label_gene = ifelse(gene %in% top15, gene, NA))

    p1 =
  res_tbl %>%
    filter(!is.na(padj)) %>%
    ggplot(aes(x = log2FoldChange, y = -log10(padj),
               color = padj < 0.05 & abs(log2FoldChange) > 0.5, 
               label = label_gene)) +
    scale_colour_manual(values = c("gray", "firebrick3")) +
    labs(col="Significantly expressed")+
    geom_point(size = 0.5) +
    geom_hline(yintercept = -log10(0.05)) +
    geom_vline(xintercept = 0.5) +
    geom_vline(xintercept = -0.5)+
    geom_text_repel(max.overlaps = 30,
                    box.padding = 0.25,
                    segment.color = 'grey50',
                    fontface = "italic")+
    labs(title = name)+
    theme_bw()
  
  print(p1)
  
    ggsave(paste0("./figures/",name,"/volcanoplot.svg"), last_plot(), 
           dpi= 500, width = 8, height =4 )
    rm(top15)

}

```



```{r, echo=FALSE}

res_tbl = read.csv("./data_output/Combined_Day1_Ola1_vs_Day1_DMSO/res_tbl.csv")

top15 = 
  res_tbl %>% 
  arrange(desc(abs(log2FoldChange))) %>% 
  filter(!is.na(gene)&padj<0.05)
    
top15 = top15$gene[1:15]

res_tbl <- res_tbl %>%
  filter(!is.na(padj)&padj<0.05) %>%
  mutate(label_gene = ifelse(gene %in% top15, gene, NA))


res_tbl %>%
    filter(!is.na(padj)) %>%
    ggplot(aes(x = log2FoldChange, y = -log10(padj),
               color = padj < 0.05 & abs(log2FoldChange) > 0.5, 
               label = label_gene)) +
    scale_colour_manual(values = c("gray", "firebrick3")) +
    labs(col="Significantly expressed")+
    geom_point(size = 0.5) +
    geom_hline(yintercept = -log10(0.05)) +
    geom_vline(xintercept = 0.5) +
    geom_vline(xintercept = -0.5)+
    geom_text_repel(max.overlaps = 100,
                    box.padding = 1.5,
                    segment.color = 'grey50',
                    fontface = "italic",
                    size = 2)+
    labs(title = "Ola1 day1 vs DMSO")+
  scale_x_continuous(limits = c(-15,25))+
    theme_bw()+ 
  theme(legend.position="none")
  

ggsave("./figures/for_pub/Ola1_Day1_volcanoplot.svg", last_plot(), 
           dpi= 300, width = 4, height = 3, units = "in")



## Ola 5

res_tbl = read.csv("./data_output/Combined_Day1_Ola5_vs_Day1_DMSO/res_tbl.csv")

top15 = 
  res_tbl %>% 
  arrange(desc(abs(log2FoldChange))) %>% 
  filter(!is.na(gene)&padj<0.05)
    
top15 = top15$gene[1:15]

res_tbl <- res_tbl %>%
  filter(!is.na(padj)&padj<0.05) %>%
  mutate(label_gene = ifelse(gene %in% top15, gene, NA))

res_tbl %>%
    filter(!is.na(padj)) %>%
    ggplot(aes(x = log2FoldChange, y = -log10(padj),
               color = padj < 0.05 & abs(log2FoldChange) > 0.5, 
               label = label_gene)) +
    scale_colour_manual(values = c("gray", "firebrick3")) +
    labs(col="Significantly expressed")+
    geom_point(size = 0.5) +
    geom_hline(yintercept = -log10(0.05)) +
    geom_vline(xintercept = 0.5) +
    geom_vline(xintercept = -0.5)+
    geom_text_repel(max.overlaps = 100,
                    box.padding = 1.5,
                    segment.color = 'grey50',
                    fontface = "italic",
                    size = 2)+
    labs(title = "Ola5 day1 vs DMSO")+
  scale_x_continuous(limits = c(-25,25))+
    theme_bw()+ 
  theme(legend.position="none")
  

 
ggsave("./figures/for_pub/Ola5_Day1_volcanoplot.svg", last_plot(), 
           dpi= 300, width = 4, height = 3, units = "in")


```


## GSEA Day1 Ola1
### GOBP

```{r, fig.height=5, fig.width=7}

res_tbl = read.csv(paste0("./data_output/Combined_Day1_Ola1_vs_Day1_DMSO/res_tbl.csv"))
    
ordered_genes_fc = res_tbl$log2FoldChange
names(ordered_genes_fc) = res_tbl$gene
ordered_genes_fc = sort(ordered_genes_fc, decreasing = T)

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_GOBP_sets,
    gs_name,
     gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 8)+ 
  labs(title = "GSEA",
          subtitle = paste0("GSEA GO:BP Day 1 Ola 1 vs DMSO"))+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

ggsave(plot = last_plot(), paste0("./figures/Combined_Day1_Ola1_vs_Day1_DMSO/GOBPgsea.svg"), dpi = 300,
      height = 8, width = 8, device = "svg")


ggsave(plot = last_plot(), paste0("./figures/for_pub/Ola1_day1_GOBPgsea.svg"), dpi = 300,
      height = 3.5, width = 6, device = "svg")

write.xlsx(gsea_results, paste0("./data_output/Combined_Day1_Ola1_vs_Day1_DMSO/GOBPgsea.xlsx"))

gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```

### Reactome

no results

### Hallmark

```{r, fig.height=5, fig.width=7}

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_hallmark_sets,
    gs_name,
     gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 8)+ 
  labs(title = "GSEA",
          subtitle = paste0("GSEA Hallmark Day 1 Ola 1 vs DMSO"))+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

ggsave(plot = last_plot(), paste0("./figures/Combined_Day1_Ola1_vs_Day1_DMSO/Hallmarkgsea.svg"), dpi = 300,
      height = 8, width = 8, device = "svg")

ggsave(plot = last_plot(), paste0("./figures/for_pub/Ola1_day1_Hallmarkgsea.svg"), dpi = 300,
      height = 2.5, width = 5, device = "svg")

write.xlsx(gsea_results, paste0("./data_output/Combined_Day1_Ola1_vs_Day1_DMSO/Hallmarkgsea.xlsx"))

gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```

### GSEA plot 

```{r}

# hallmark p53 pathway

gseaplot2(gsea_results, geneSetID = 2, title = gsea_results$Description[2])

ggsave(plot = last_plot(), paste0("./figures/for_pub/Ola1_day1_gseaplot_p53.svg"), dpi = 300,
      height = 2.5, width = 4, device = "svg")

```


## GSEA Day1 Ola5
### GOBP

```{r}

res_tbl = read.csv(paste0("./data_output/Combined_Day1_Ola5_vs_Day1_DMSO/res_tbl.csv"))
    
ordered_genes_fc = res_tbl$log2FoldChange
names(ordered_genes_fc) = res_tbl$gene
ordered_genes_fc = sort(ordered_genes_fc, decreasing = T)

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_GOBP_sets,
    gs_name,
     gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 8)+ 
  labs(title = "GSEA",
          subtitle = paste0("GSEA GO:BP Day 1 Ola 5 vs DMSO"))+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

ggsave(plot = last_plot(), paste0("./figures/Combined_Day1_Ola5_vs_Day1_DMSO/GOBPgsea.svg"), dpi = 300,
      height = 8, width = 8, device = "svg")

ggsave(plot = last_plot(), paste0("./figures/for_pub/Ola5_day1_GOBPgsea.svg"), dpi = 300,
      height = 5, width = 5.5, device = "svg")

write.xlsx(gsea_results, paste0("./data_output/Combined_Day1_Ola5_vs_Day1_DMSO/GOBPgsea.xlsx"))

gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```

### Reactome

```{r}

res_tbl = read.csv("./data_output/Combined_Day1_Ola5_vs_Day1_DMSO/res_tbl.csv")
    
ordered_genes_fc = res_tbl$log2FoldChange
names(ordered_genes_fc) = res_tbl$gene
ordered_genes_fc = sort(ordered_genes_fc, decreasing = T)

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_reactome_sets,
    gs_name,
     gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 8)+ 
  labs(title = "GSEA",
          subtitle = paste0("GSEA Reactome Day 1 Ola 5 vs DMSO"))+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

ggsave(plot = last_plot(), paste0("./figures/Combined_Day1_Ola5_vs_Day1_DMSO/Reactomegsea.svg"), dpi = 300,
      height = 8, width = 8, device = "svg")
ggsave(plot = last_plot(), paste0("./figures/for_pub/Ola5_day1_Reactomegsea.svg"), dpi = 300,
      height = 6, width = 5.5, device = "svg")

write.xlsx(gsea_results, paste0("./data_output/Combined_Day1_Ola5_vs_Day1_DMSO/Reactomegsea.xlsx"))

gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```

### Hallmark

```{r, fig.height=5, fig.width=7}

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_hallmark_sets,
    gs_name,
     gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 8)+ 
  labs(title = "GSEA",
          subtitle = paste0("GSEA Hallmark Day 1 Ola 5 vs DMSO"))+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

ggsave(plot = last_plot(), paste0("./figures/Combined_Day1_Ola5_vs_Day1_DMSO/Hallmarkgsea.svg"), dpi = 300,
      height = 8, width = 8, device = "svg")
ggsave(plot = last_plot(), paste0("./figures/for_pub/Ola5_day1_Hallmarkgsea.svg"), dpi = 300,
      height = 3, width = 5.5, device = "svg")

write.xlsx(gsea_results, paste0("./data_output/Combined_Day1_Ola5_vs_Day1_DMSO/Hallmarkgsea.xlsx"))

gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```

### GSEA plot 

```{r}

# hallmark p53 pathway

gseaplot2(gsea_results, geneSetID = 3, title = gsea_results$Description[3])

ggsave(plot = last_plot(), paste0("./figures/for_pub/Ola5_day1_gseaplot_p53.svg"), dpi = 300,
      height = 2.5, width = 4, device = "svg")

```

### Summary

```{r}

# shared with ola 1 are
# GOBP_NCRNA_PROCESSING GOBP_RIBOSOME_BIOGENESIS
# HALLMARK_P53_PATHWAY HALLMARK_MYC_TARGETS_V1

# pool terms of interest
hsa_summary = 
  hsa_GOBP_sets %>% 
  filter(gs_name %in% toupper(c("gobp_ribosome_biogenesis",
                        "gobp_ncrna_processing","GOBP_RRNA_METABOLIC_PROCESS",
                        "GOBP_TRNA_METABOLIC_PROCESS", "GOBP_RIBONUCLEOPROTEIN_COMPLEX_BIOGENESIS",
                        "GOBP_CHROMATIN_ASSEMBLY_OR_DISASSEMBLY	"))) %>% 
  rbind(hsa_reactome_sets %>% filter(
    gs_name%in% toupper(c("REACTOME_CELL_CYCLE_CHECKPOINTS",
                          "REACTOME_G2_M_CHECKPOINTS",
                          "REACTOME_TRANSLATION",
                          "REACTOME_CHROMOSOME_MAINTENANCE",
                          "REACTOME_ESTROGEN_DEPENDENT_GENE_EXPRESSION",
                          "REACTOME_SENESCENCE_ASSOCIATED_SECRETORY_PHENOTYPE_SASP",
                          "REACTOME_HDACS_DEACETYLATE_HISTONES")))) %>% 
  rbind(hsa_hallmark_sets %>%  filter(gs_name %in% toupper(c("hallmark_p53_pathway",
                                                            "hallmark_myc_targets_v1",
                                                            "HALLMARK_E2F_TARGETS"))))

# load results for each conditions
res_tbl_ola1 = read.csv("./data_output/Combined_Day1_Ola1_vs_Day1_DMSO/res_tbl.csv")
res_tbl_ola5 = read.csv("./data_output/Combined_Day1_Ola5_vs_Day1_DMSO/res_tbl.csv")


ordered_genes_fc1 <- res_tbl_ola1$log2FoldChange
names(ordered_genes_fc1) <- res_tbl_ola1$gene
ordered_genes_fc1 <- sort(ordered_genes_fc1, decreasing = T)
table(is.na(ordered_genes_fc1))

ordered_genes_fc2 <- res_tbl_ola5$log2FoldChange
names(ordered_genes_fc2) <- res_tbl_ola5$gene
ordered_genes_fc2 <- sort(ordered_genes_fc2, decreasing = T)

gcSample = list(Ola_1 = ordered_genes_fc1, 
                Ola_5 = ordered_genes_fc2)

# compare

ck = compareCluster(gcSample, fun = GSEA, TERM2GENE = dplyr::select(
    hsa_summary,
    gs_name,
    gene_symbol))

# plot

dotplot(ck, showCategory = 20, font.size = 7, x = "NES", by = "Count")+
 theme(strip.background = element_rect(
     color="black", fill = 'grey95',size=.5, linetype="solid"))+
  facet_grid(.~Cluster)+
  scale_colour_gradient2(limits=c(0, 0.05), low="red", high = "blue", 
                         mid = "grey", midpoint = 0.015)+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank())+
  labs(title = "Day 1 GSEA",
              subtitle = "Ola 1 and Ola 5 compared to DMSO")

ggsave(plot = last_plot(), paste0("./figures/for_pub/Summary_day1_gsea.svg"), dpi = 300,
      height = 5, width = 8, device = "svg")

```


# Day 5 Ola 1 and 5 vs DMSO

## Volcano plots

```{r, fig.show="hold", out.width="50%"}

for (i in c(5:6)){
    cond = resultsNames(dds5)[i]
    name = gsub("Treatment_","",resultsNames(dds5)[i])
    
    res <- results(dds5, name = cond)
    res_tbl <- as_tibble(res, rownames="ensembl")
    res_tbl <- res_tbl %>%
    left_join(ensembl_to_geneName) %>%
    arrange(padj) 
    dir.create(paste0(dir,'/data_output/',name))
    dir.create(paste0(dir,'/figures/',name))
    write.csv(res_tbl, paste0("./data_output/",name,"/res_tbl.csv"), row.names = T)
    overexpr <- res_tbl %>% filter(padj<=0.05 & log2FoldChange>=0.5) 
    underexpr <- res_tbl %>% filter(padj<=0.05 & log2FoldChange<=-0.5)
    signif <- full_join(overexpr, underexpr)
  
    write.xlsx(overexpr, file= paste0("./data_output/",name,"/Sign_genes.xlsx"), sheetName="UP")
    write.xlsx(underexpr, file=paste0("./data_output/",name,"/Sign_genes.xlsx"), sheetName="DOWN", append=TRUE)
    write.xlsx(signif, file=paste0("./data_output/",name,"/Sign_genes.xlsx"), sheetName="SIGNIF", append=TRUE)
    
    p1 =
  res_tbl %>%
    filter(!is.na(padj)) %>%
    ggplot(aes(x = log2FoldChange, y = -log10(padj),
               color = padj < 0.05 & abs(log2FoldChange) > 1, 
               label = ifelse(padj<0.05&log2FoldChange>=1|
                                padj<0.05&log2FoldChange<=-1,as.character(gene),''))) +
    scale_colour_manual(values = c("gray", "firebrick3")) +
    labs(col="Significantly expressed")+
    geom_point(size = 0.5) +
    geom_hline(yintercept = -log10(0.05)) +
    geom_vline(xintercept = 1) +
    geom_vline(xintercept = -1)+
    geom_text_repel(max.overlaps = 15,
                    box.padding = 0.25,
                    segment.color = 'grey50',
                    fontface = "italic")+
    labs(title = name)+
    theme_bw()
  
  print(p1)
  
    ggsave(paste0("./figures/",name,"/volcanoplot.svg"), last_plot(), 
           dpi= 300, width = 12, height = 8)
  
    
}


```

```{r, echo=FALSE}

res_tbl = read.csv("./data_output/Combined_Day5_Ola1_vs_Day5_DMSO/res_tbl.csv")

top15 = 
  res_tbl %>% 
  arrange(desc(abs(log2FoldChange))) %>% 
  filter(!is.na(gene)&padj<0.05)
    
top15 = top15$gene[1:15]

res_tbl <- res_tbl %>%
  filter(!is.na(padj)&padj<0.05) %>%
  mutate(label_gene = ifelse(gene %in% top15, gene, NA))


res_tbl %>%
    filter(!is.na(padj)) %>%
    ggplot(aes(x = log2FoldChange, y = -log10(padj),
               color = padj < 0.05 & abs(log2FoldChange) > 0.5, 
               label = label_gene) ) +
    scale_colour_manual(values = c("gray", "firebrick3")) +
    labs(col="Significantly expressed")+
    geom_point(size = 0.5) +
    geom_hline(yintercept = -log10(0.05)) +
    geom_vline(xintercept = 0.5) +
    geom_vline(xintercept = -0.5)+
    geom_text_repel(max.overlaps = 50,
                    box.padding = 1.5,
                    segment.color = 'grey50',
                    fontface = "italic",
                    size = 2)+
    labs(title = "Ola1 day5 vs DMSO")+
  scale_x_continuous(limits = c(-15,25))+
    theme_bw()+ 
  theme(legend.position="none")
  

 
ggsave("./figures/for_pub/Ola1_Day5_volcanoplot.svg", last_plot(), 
           dpi= 300, width = 4, height = 3, units = "in")


## Ola 5

res_tbl = read.csv("./data_output/Combined_Day5_Ola5_vs_Day5_DMSO/res_tbl.csv")

top15 = 
  res_tbl %>% 
  arrange(desc(abs(log2FoldChange))) %>% 
  filter(!is.na(gene)&padj<0.05)
    
top15 = top15$gene[1:15]

res_tbl <- res_tbl %>%
  filter(!is.na(padj)&padj<0.05) %>%
  mutate(label_gene = ifelse(gene %in% top15, gene, NA))


res_tbl %>%
    filter(!is.na(padj)) %>%
    ggplot(aes(x = log2FoldChange, y = -log10(padj),
               color = padj < 0.05 & abs(log2FoldChange) > 0.5, 
               label = label_gene) ) +
    scale_colour_manual(values = c("gray", "firebrick3")) +
    labs(col="Significantly expressed")+
    geom_point(size = 0.5) +
    geom_hline(yintercept = -log10(0.05)) +
    geom_vline(xintercept = 0.5) +
    geom_vline(xintercept = -0.5)+
    geom_text_repel(max.overlaps = 50,
                    box.padding = 1.5,
                    segment.color = 'grey50',
                    fontface = "italic",
                    size = 2)+
    labs(title = "Ola5 day5 vs DMSO")+
  scale_x_continuous(limits = c(-15,25))+
    theme_bw()+ 
  theme(legend.position="none")
  

 
ggsave("./figures/for_pub/Ola5_Day5_volcanoplot.svg", last_plot(), 
           dpi= 300, width = 4, height = 3, units = "in")


```


## GSEA Day5 Ola1
### GOBP

```{r}

res_tbl = read.csv(paste0("./data_output/Combined_Day5_Ola1_vs_Day5_DMSO/res_tbl.csv"))
    
ordered_genes_fc = res_tbl$log2FoldChange
names(ordered_genes_fc) = res_tbl$gene
ordered_genes_fc = sort(ordered_genes_fc, decreasing = T)

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_GOBP_sets,
    gs_name,
     gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 6)+ 
  labs(title = "GSEA",
          subtitle = paste0("GSEA GO:BP Day 5 Ola 1 vs DMSO"))+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

ggsave(plot = last_plot(), paste0("./figures/Combined_Day5_Ola1_vs_Day5_DMSO/GOBPgsea.svg"), dpi = 300,
      height = 8, width = 8, device = "svg")

ggsave(plot = last_plot(), paste0("./figures/for_pub/Ola1_day5_GOBPgsea.svg"), dpi = 300,
      height = 5, width = 4.5, device = "svg")

write.xlsx(gsea_results, paste0("./data_output/Combined_Day5_Ola1_vs_Day5_DMSO/GOBPgsea.xlsx"))

gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```


### Reactome

```{r}

res_tbl = read.csv(paste0("./data_output/Combined_Day5_Ola1_vs_Day5_DMSO/res_tbl.csv"))
    
ordered_genes_fc = res_tbl$log2FoldChange
names(ordered_genes_fc) = res_tbl$gene
ordered_genes_fc = sort(ordered_genes_fc, decreasing = T)

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_reactome_sets,
    gs_name,
     gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 8)+ 
  labs(title = "GSEA",
          subtitle = paste0("GSEA Reactome Day 5 Ola 1 vs DMSO"))+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

ggsave(plot = last_plot(), paste0("./figures/Combined_Day5_Ola1_vs_Day5_DMSO/Reactomegsea.svg"), dpi = 300,
      height = 8, width = 8, device = "svg")
ggsave(plot = last_plot(), paste0("./figures/for_pub/Ola1_day5_Reactomegsea.svg"), dpi = 300,
      height = 5, width = 5, device = "svg")

write.xlsx(gsea_results, paste0("./data_output/Combined_Day5_Ola1_vs_Day5_DMSO/Reactomegsea.xlsx"))

gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```


### GSEA plot 

```{r}

# fatty acid

gseaplot2(gsea_results, geneSetID = 21, title = gsea_results$Description[21])

ggsave(plot = last_plot(), paste0("./figures/for_pub/Ola1_day5_gseaplot_fattyacid.svg"), dpi = 300,
      height = 2.5, width = 4, device = "svg")

```


### Hallmark

```{r, fig.height=5, fig.width=7}

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_hallmark_sets,
    gs_name,
     gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 8)+ 
  labs(title = "GSEA",
          subtitle = paste0("GSEA Hallmark Day 5 Ola 1 vs DMSO"))+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

ggsave(plot = last_plot(), paste0("./figures/Combined_Day5_Ola1_vs_Day5_DMSO/Hallmarkgsea.svg"), dpi = 300,
      height = 8, width = 8, device = "svg")
ggsave(plot = last_plot(), paste0("./figures/for_pub/Ola1_day5_Hallmarkgsea.svg"), dpi = 300,
      height = 5, width = 5.5, device = "svg")

write.xlsx(gsea_results, paste0("./data_output/Combined_Day5_Ola1_vs_Day5_DMSO/Hallmarkgsea.xlsx"))

gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")

```

### GSEA plot 

```{r}

# hallmark emt pathway

gseaplot2(gsea_results, geneSetID = 12, title = gsea_results$Description[12])

ggsave(plot = last_plot(), paste0("./figures/for_pub/Ola1_day5_gseaplot_emt.svg"), dpi = 300,
      height = 2.5, width = 4, device = "svg")

```

## GSEA Day5 Ola5
### GOBP

```{r}

res_tbl = read.csv(paste0("./data_output/Combined_Day5_Ola5_vs_Day5_DMSO/res_tbl.csv"))
    
ordered_genes_fc = res_tbl$log2FoldChange
names(ordered_genes_fc) = res_tbl$gene
ordered_genes_fc = sort(ordered_genes_fc, decreasing = T)

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_GOBP_sets,
    gs_name,
     gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 8)+ 
  labs(title = "GSEA",
          subtitle = paste0("GSEA GO:BP Day 5 Ola 5 vs DMSO"))+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

ggsave(plot = last_plot(), paste0("./figures/Combined_Day5_Ola5_vs_Day5_DMSO/GOBPgsea.svg"), dpi = 300,
      height = 8, width = 8, device = "svg")
ggsave(plot = last_plot(), paste0("./figures/for_pub/Ola5_day5_GOBPgsea.svg"), dpi = 300,
      height = 5, width = 5.5, device = "svg")

write.xlsx(gsea_results, paste0("./data_output/Combined_Day5_Ola5_vs_Day5_DMSO/GOBPgsea.xlsx"))

gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```

### Reactome

```{r}

res_tbl = read.csv(paste0("./data_output/Combined_Day5_Ola5_vs_Day5_DMSO/res_tbl.csv"))
    
ordered_genes_fc = res_tbl$log2FoldChange
names(ordered_genes_fc) = res_tbl$gene
ordered_genes_fc = sort(ordered_genes_fc, decreasing = T)

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_reactome_sets,
    gs_name,
     gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 6)+ 
  labs(title = "GSEA",
          subtitle = paste0("GSEA Reactome Day 5 Ola 5 vs DMSO"))+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

ggsave(plot = last_plot(), paste0("./figures/Combined_Day5_Ola5_vs_Day5_DMSO/Reactomegsea.svg"), dpi = 300,
      height = 8, width = 8, device = "svg")
ggsave(plot = last_plot(), paste0("./figures/for_pub/Ola5_day5_Reactomegsea.svg"), dpi = 300,
      height = 5, width = 4.5, device = "svg")

write.xlsx(gsea_results, paste0("./data_output/Combined_Day5_Ola5_vs_Day5_DMSO/Reactomegsea.xlsx"))

gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```

### GSEA plot 

```{r}

# fatty acid

gseaplot2(gsea_results, geneSetID = 82, title = gsea_results$Description[82])

ggsave(plot = last_plot(), paste0("./figures/for_pub/Ola5_day5_gseaplot_fattyacid.svg"), dpi = 300,
      height = 2.5, width = 4, device = "svg")


```


### Hallmark

```{r, fig.height=5, fig.width=7}

gsea_results = GSEA(
  geneList = ordered_genes_fc, 
  pvalueCutoff = 0.05, 
  eps = 0, 
  pAdjustMethod = "BH", 
  nPermSimple = 10000,
  TERM2GENE = dplyr::select(
    hsa_hallmark_sets,
    gs_name,
     gene_symbol
  )
)

dotplot(gsea_results, x = "NES", showCategory = 20, font.size = 6)+ 
  labs(title = "GSEA",
          subtitle = paste0("GSEA Hallmark Day 5 Ola 5 vs DMSO"))+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

ggsave(plot = last_plot(), paste0("./figures/Combined_Day5_Ola1_vs_Day5_DMSO/Hallmarkgsea.svg"), dpi = 300,
      height = 8, width = 8, device = "svg")
ggsave(plot = last_plot(), paste0("./figures/for_pub/Ola5_day5_Hallmarkgsea.svg"), dpi = 300,
      height = 5, width = 4.5, device = "svg")

write.xlsx(gsea_results, paste0("./data_output/Combined_Day5_Ola1_vs_Day5_DMSO/Hallmarkgsea.xlsx"))

gsea_results %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  select(ID, NES, setSize,	p.adjust,	qvalue, core_enrichment) %>%
  knitr::kable()%>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width = "1000px", height = "500px")
```

### GSEA plot 

```{r}

# hallmark emt pathway

gseaplot2(gsea_results, geneSetID = 10, title = gsea_results$Description[10])

ggsave(plot = last_plot(), paste0("./figures/for_pub/Ola5_day5_gseaplot_emt.svg"), dpi = 300,
      height = 2.5, width = 4, device = "svg")


```

### Summary

```{r}

hsa_summary = 
  hsa_GOBP_sets %>% 
  filter(gs_name %in% toupper(c("GOBP_DOUBLE_STRAND_BREAK_REPAIR", "GOBP_LIPID_CATABOLIC_PROCESS","GOBP_FATTY_ACID_METABOLIC_PROCESS","GOBP_NEGATIVE_REGULATION_OF_IMMUNE_SYSTEM_PROCESS", "GOBP_STEROL_METABOLIC_PROCESS","GOBP_RESPONSE_TO_INTERFERON_GAMMA","GOBP_MACROPHAGE_ACTIVATION","GOBP_RESPONSE_TO_INTERLEUKIN_1","GOBP_NEUTROPHIL_CHEMOTAXIS","GOBP_T_CELL_MIGRATION"))) %>% 
  rbind(hsa_reactome_sets %>% filter(
    gs_name%in% toupper(c("REACTOME_FATTY_ACID_METABOLISM", "REACTOME_CELL_CYCLE_CHECKPOINTS",
                          "REACTOME_ECM_PROTEOGLYCANS", "GOBP_INTERLEUKIN_2_PRODUCTION","GOBP_REGULATION_OF_SPROUTING_ANGIOGENESIS","GOBP_FAT_SOLUBLE_VITAMIN_CATABOLIC_PROCESS","GOBP_TRANSFORMING_GROWTH_FACTOR_BETA_PRODUCTION","GOBP_REGULATION_OF_MEMBRANE_LIPID_DISTRIBUTION")))) %>% 
  rbind(hsa_hallmark_sets %>%  filter(gs_name %in% toupper(c("hallmark_p53_pathway","HALLMARK_TNFA_SIGNALING_VIA_NFKB","HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION", "hallmark_hypoxia","HALLMARK_UNFOLDED_PROTEIN_RESPONSE","HALLMARK_MTORC1_SIGNALING","HALLMARK_INFLAMMATORY_RESPONSE"))))

# load results for each conditions
res_tbl_ola1 = read.csv("./data_output/Combined_Day5_Ola1_vs_Day5_DMSO/res_tbl.csv")
res_tbl_ola5 = read.csv("./data_output/Combined_Day5_Ola5_vs_Day5_DMSO/res_tbl.csv")


ordered_genes_fc1 <- res_tbl_ola1$log2FoldChange
names(ordered_genes_fc1) <- res_tbl_ola1$gene
ordered_genes_fc1 <- sort(ordered_genes_fc1, decreasing = T) 

ordered_genes_fc2 <- res_tbl_ola5$log2FoldChange
names(ordered_genes_fc2) <- res_tbl_ola5$gene
ordered_genes_fc2 <- sort(ordered_genes_fc2, decreasing = T)

gcSample1 = list(Ola_1 = ordered_genes_fc1, 
                Ola_5 = ordered_genes_fc2)

# compare

ck = compareCluster(gcSample1, fun = GSEA, TERM2GENE = dplyr::select(
    hsa_summary,gs_name,gene_symbol))

# plot

dotplot(ck, showCategory = 20, font.size = 7, x = "NES", by = "Count")+
 theme(strip.background = element_rect(
     color="black", fill = 'grey95',size=.5, linetype="solid"))+
  facet_grid(.~Cluster)+
  scale_colour_gradient2(limits=c(0, 0.05), low="red", high = "blue", 
                         mid = "grey", midpoint = 0.015)+
  theme(axis.line = element_line(color='black'),
  plot.background = element_blank())+
  labs(title = "Day 5 GSEA",
              subtitle = "Ola 1 and Ola 5 compared to DMSO")

ggsave(plot = last_plot(), paste0("./figures/for_pub/Summary_day5_gsea.svg"), dpi = 300,
      height = 6, width = 8, device = "svg")

```

# Venn diagram on steroids (UpsetPlot)

Conditions we want to compare, per group:

\- Ola 1 vs DMSO

\- Ola 5 vs DMSO

I extracted all significantly differentially expressed genes (DEGs) for
each condition (\|log2Fc\| \>= 1 and pval \< 0.05).

**Set size** = how many DEGs there are per sample

**Intersection size** = How many genes are shared between the x samples
tied together with the bar + nodes

So if you sum up all the intersection size with a dot in a sample, you
get the set size of that sample. It's essentially a venn diagram for
when you have \>3 samples you want to compare.

```{r, fig.height=7, fig.width=10}

# load in the data to be formatted for an upset plot
# keep each gene name that has a signif pval and |log2fc| > 0.5
# use ensembl to encompass unnamed transcripts

folder = c("Combined_Day1_Ola1_vs_Day1_DMSO","Combined_Day1_Ola5_vs_Day1_DMSO",
           "Combined_Day5_Ola1_vs_Day5_DMSO","Combined_Day5_Ola5_vs_Day5_DMSO")

for(i in 1:length(folder)){
  res = read.csv(paste0("./data_output/",folder[i],"/res_tbl.csv"))
  genes = res %>% 
    select(ensembl, log2FoldChange, padj) %>% 
    filter(abs(log2FoldChange)>= 0.5 & padj < 0.05) %>% 
    select(ensembl)
  genes$temp = 1
  colnames(genes)[2] = folder[i]
  assign(folder[i], genes)
}

datalist_dmso = list(Combined_Day1_Ola1_vs_Day1_DMSO,Combined_Day1_Ola5_vs_Day1_DMSO,
           Combined_Day5_Ola1_vs_Day5_DMSO,Combined_Day5_Ola5_vs_Day5_DMSO)

big_data_dmso =
  datalist_dmso %>% purrr::reduce(full_join, by='ensembl')%>% 
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) 

```

```{r, fig.height=5, fig.width=10}

m1 = make_comb_mat(big_data_dmso)

UpSet(m1, top_annotation = upset_top_annotation(m1, add_numbers = TRUE),
    right_annotation = upset_right_annotation(m1, add_numbers = TRUE),
    comb_order = order(-comb_size(m1)), 
    comb_col = c("#ffbe4f", "#e8702a", "#6bd2db","#7e073d")[comb_degree(m1)],
    row_names_gp = gpar(fontsize = 10))


```

